---
output:
  word_document: default
  html_document: default
---
title: "Cyclistic Case Study"
author: "Rebecca Green"
date: "2024-05-29"
output:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
This case study is part of Google Data Analytics on Coursera. In the case study scenario, you are casted as a junior data analyst on the marketing team at Cyclistic. Cyclistic is a fictional bike-share company in Chicago. There is a question you have to answer using data and data visualization.

## 1. "Ask"

1. How do annual members and casual riders use Cyclistic bikes differently?

## 2. "Prepare" 

_note__: All the data preparation processes and tools involved were all under **Windows 11 (Home)** operating system.

The [**data**](https://divvy-tripdata.s3.amazonaws.com/index.html) used in this scenario was provided by the course under this [**license**](https://www.divvybikes.com/data-license-agreement). There were 2 data files for the data collected. Each file was originally compressed in a .zip file format. The .zip were downloaded and extracted. 


## 3. "Process"

### 3.1 Processing Data 

### 3.1,1 Installing packages 

```{r}

#Code chunk 3.1,1.1

library(tidyverse)
library(conflicted)

```

#### 3.1,2 Uploading data sets 

```{r}

#Code chunk 3.1.2,1 

q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")

```

#### 3.1.3 Comparing columns and renaming them for consistency 

```{r}

#Code chunk 3.1.3.1 

colnames(q1_2019)
colnames(q1_2020)

```

```{r}

#Code chunk 3.1,2.1 

(q1_2019 <- rename(q1_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype
))

```

#### 3.1.4  Inspecting the dataframes and look for incongruencies and converting ride_id and rideable_type to character so that they can stack correctly

```{r}

#Code chunk 3.1.4.1

str(q1_2019)
str(q1_2020)

```

```{r}

#Code chunk 3.1,4,2

q1_2019 <-  mutate(q1_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```

#### 3.1.5 Stacking data frames into one and removing no longer used fields 

```{r}

#Code chunk 3.1,5.1

all_trips <- bind_rows(q1_2019, q1_2020)#, q3_2019)#, q4_2019, q1_2020)

```

```{r}

#Code chunk 3.1.5.2 

all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender,  "tripduration"))

```

#### 3.1.6 Inspecting the new table and reassigning the desired values(we will go with the current 2020 labels) and checking to make sure the proper number of observations were reassigned

```{r}

#Code chunk 3.1.6.1 

colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics

```

```{r}

#Code chunk 3.1.6.2

table(all_trips$member_casual)
      
```

```{r}

#Code chunk 3.1.6.3 

all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

```

```{r}

#Code chunk 3.1,6.4

table(all_trips$member_casual)

```

#### 3.1.7 Adding columns that list the date, month, day, and year of each ride and adding a "ride_length" calculation to all_trips (in seconds)

```{r}

#Code chunk 3.1.7.1 

all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

```{r}

#Code chunk 3.1.7.2

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

```

#### 3.1.8 Inspecting columns and converting "ride_length" from Factor to numeric so we can run calculations on the data and removing bad data 

```{r}

#Code chunk 3.1.8.1 

str(all_trips)

```

```{r}

#Code chunk 3.1.8.2

is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```

```{r}

#Code chunk 3.1.8.3 

all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```

## 4. "Analyze"

### 4.1 Analying data 

#### 4.1.1 Analying data on ride_length (all figures in seconds)

```{r}

#Code chunk 4.1.1

summary(all_trips_v2$ride_length)

```

#### 4.1.2 Comparing members and casual users and seeing the average ride time by each day for members vs casual users

```{r}

#Code chunk 4.1.2.1

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

```{r}

#Code chunk 4.1.2.2

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

#### 4.1.3 Fixing the days of the week and running the average ride time by each day for members vs casual users and visualizing data 

```{r}

#Code chunk 4.1.3.1

all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

```{r}

#Code chunk 4.1.3.2

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

```{r}

#Code chunk 4.1.3.3

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts


```

```{r}

#Code chunk 4.1.3.4

all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n(),
            average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +
  scale_y_continuous(name = "number_of_rides", labels = scales::comma)


```

```{r}

#Code chunk 4.1.3.5

all_trips_v2 %>% 
        mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")


```

#### 4.2 What the data visualization shows 

1. Annual Members take more rides during the time frame 

2. Casual members ride longer rides than annual members. 
`


## 5. "Share"

I will share this on my Github account:(https://github.com/rg0730/Coursera-Cyclistic-Case-Study)

## 6. "Act"

### 6.1 Summary and insight

From the two plots, it is clear that the Annual members take more rides. This could be because, they use the Cycvlistic bikes in their day to day life. On the other hand, it is clear that the casual members ride longer rides than annual members. This could mean that they are tourists. But both of these could be statements need more data to see if they are correct.

### 6.2 Suggestions

#### 6.2.1 Gather more data about customers 

This gathering more data about the customers home addresses would allow to company to see if some of the casual members are tourists 

#### 6.2.2 Expand the branches 

This would convert some casual members to annual members if they could use the bikes in more than just Chicago.

#### 6.3 3. Add more annual member exclusive benefits 

If there was more exclusive benefits to being an annual member like being able to reserve parking spaces. Or another one could be extended riding times without extra fees. Lastly, one could be priority access to bikes during busy times or at busy stations. 